\documentclass[12pt, titlepage]{article}
%Packages
\usepackage{
  acronym,
  enumerate,
  fancyhdr,
  hyperref,
  indentfirst,
  lastpage,
  listings,
  microtype,
}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[htt]{hyphenat}

% declare some useful macros
\newcommand{\assignmentNumber}{Assignment 4}
\newcommand{\courseName}{CS 444}
\newcommand{\z}[1]{\texttt{#1}}

\acrodef{jls}[JLS]{Java Language Specification}

\lstset{basicstyle=\ttfamily}

\begin{document}
\pagestyle{fancyplain}
\thispagestyle{plain}

%Headers
\fancyhead{}
\fancyfoot{}
\rhead{\fancyplain{}{Page \thepage\ of \pageref*{LastPage}}}
\chead{\fancyplain{}{\assignmentNumber}}
\lhead{\fancyplain{}{\courseName}}

\title{CS444 - \assignmentNumber\\Name Resolution, Type Checking, and Static Analysis}
\date{\today}
\author{Alex Klen\\20372654\and Sanjay Menakuru\\20374915\and Jonathan Wei\\20376489}

\maketitle

\section{Introduction}

We decided to tackle assignments 2, 3, and 4 together, as we realized that all
of the necessary computation for these three assignments could be carried out
in the same set of AST traversals. Our resulting code is structured in five
phases, described in some detail in the \nameref{sec:arch} section.

\section{Architecture}\label{sec:arch}

\subsection{Phases of the compiler}

We will consider only phases of the compiler that were added for assignment 2
through assignment 4. These phases are the following:
\begin{enumerate}
  \item \nameref{subsubsec:col-types}
  \item \nameref{subsubsec:fields}
  \item \nameref{subsubsec:type-checking}
  \item \nameref{subsubsec:const-prop}
  \item \nameref{subsubsec:data-flow}
\end{enumerate}

Here is a mapping between the various sections of the assignments and the 5
phases above:
\begin{enumerate}[\bf{A}1\ \ ]
  \setcounter{enumi}{1}

  \item
  \begin{description}
    \item[Environment Building]
      \hyperref[subsubsec:col-types]{Phase 1}, \hyperref[subsubsec:fields]{phase 2},
      and \hyperref[subsubsec:type-checking]{phase 3}
    \item[Type Linking]
      \hyperref[subsubsec:col-types]{Phase 1} and \hyperref[subsubsec:fields]{phase 2}
    \item[Hierarchy Checking]
      \hyperref[subsubsec:fields]{Phase 2}
  \end{description}

  \item
  \begin{description}
    \item[Name Disambiguation]
      \hyperref[subsubsec:type-checking]{Phase 3} and \hyperref[subsubsec:data-flow]{phase 5}
    \item[Type Checking]
      \hyperref[subsubsec:type-checking]{Phase 3}
  \end{description}

  \item
  \begin{description}
    \item[Reachability Checking]
      \hyperref[subsubsec:const-prop]{Phase 4} and \hyperref[subsubsec:data-flow]{phase 5}
  \end{description}
\end{enumerate}

\subsubsection{Collecting Types}\label{subsubsec:col-types}

This phase walks over all compilation units and collects all type names. This
is done in \z{types/types.cpp}. It adds all packages and types in a particular
compilation unit and adds them to a \z{TypeSet}.

A \z{TypeSet} is a data structure that maps a qualified name to a type. The
public API is in \z{types/typeset.\{h,cpp\}}, and the implementation is in
\z{types/typeset\_impl.\{h,cpp\}}. A \z{TypeSet} supports scoped-lookup; for instance,
it has a method called \z{WithImports} that provides a `view' into the \z{TypeSet}
assuming the provided imports are in scope. Other such methods include
\z{WithPackage} and \z{WithType}. As future phases recurse through the AST, they
call the appropriate methods on \z{TypeSet} to obtain the correct scoped view of
visible types.

When building a \z{TypeSet}, we verify that each qualified name refers to a
unique class, or a package, but not both. We emit an error for each violation
of this constraint, and suppress any future errors referencing this type. This
is done in the \z{TypeSetBuilder::Build} method which is located in
\z{types/typeset.cpp}.

\subsubsection{Collecting Fields and Methods}\label{subsubsec:fields}

First, this phase walks through the body of every type and records information
about all fields and methods. This information is stored in a class named
\z{TypeInfoMap}, which can be found in \z{types/type\_info\_map.\{h,cpp\}}. \\
We record each field's name, modifiers, containing class, and type. \\
We record each method's name, modifiers, parameter types, containing class,
return type, and a flag \z{is\_constructor}.\\

Next, this phase verifies that the implements-extends graph is acyclic and
well-formed. By well-formed, we mean that it verifies that no interface extends
a class, no class implements another class, and no class extends an interface.
The cycle-checker also produces a topological ordering of the types that
guarantees that for all types $T$, $T$'s parents are visited before $T$
itself.\\

Finally, this phase `pushes down' all method and fields from parents to
children. We associate with each type a \z{FieldTable} and a \z{MethodTable}. These
contain the types' members, and all inherited members.\\
When pushing down members, we validate several rules of Joos; here is a
non-exhaustive list of these rules:
\begin{enumerate}
  \item An inherited method cannot lower the visibility of a method in a parent
  class.

  \item A parent class must contain a zero-argument constructor.

  \item A class with abstract methods must be declared abstract.

  \item A class may not have multiple fields with the same name.

  \item A class may not have multiple methods with the same signature, where a method's signature is composed of its name and argument types.
\end{enumerate}

\subsubsection{Type Checking}\label{subsubsec:type-checking}

This phase validates that the program obeys the Joos typing rules. It uses the
\z{TypeSet} and the \z{TypeInfoMap} from the previous two phases. The source code
for this phase can be found in \z{types/typechecker.\{h,cpp\}}. It is defined as a
Visitor implementation that rewrites the AST to have type information
associated with all expressions and declarations.\\

This phase also introduces the \z{SymbolTable} class, which is located in
\z{types/symbol\_table.\{h,cpp\}}. This class manages a simple mapping from a string
to a type. It supports entering and leaving scopes, and enforces Joos's
restriction on variables with overlapping scopes. Specifically, it emits an error if
variables with identical names have an overlapping scope. It also disallows
variables from being referenced in their own initializers. The type checker
implements `alpha-renaming'; that is to say, it gives every variable a unique
identifier to distinguish two variables with the same name in different scopes.\\

Each node in the AST first type checks its children by visiting them. If any of
its children fail to type check or are pruned, then the node will assume that
an error was emitted below, and will prune itself from the AST without emitting
further errors; this pruning will avoid the cascade of errors commonly found in
many compilers.\\
If the node's children successfully type check, we check whether their types conform to the rules provided in the \ac{jls} for the appropriate expression or statement. If the rules are not met we emit an error and begin pruning this subtree. If the rules are met, we rewrite the node with its computed type information. \\

All ambiguous qualified names are handled using the algorithm provided in the
\ac{jls}. Specifically, we first try to resolve the first segment of a
qualified name in the \z{SymbolTable}. Next, we try and parse the first segment
as a field reference in the containing class's \z{FieldTable}. Finally, we try to
match a prefix of the qualified name in the \z{TypeSet}. If we find a match, we
split the qualified name into different AST types that are more specific to the
usage. For instance, we rewrite a local field reference to a \z{FieldDerefExpr}.

The same rules are also applied to all \z{CallExpr}s, with a slight modification.
We simply slice off the last part of the qualified name, and recurse on the
remaining parts. Once the remaining parts have been resolved, we then look for
a method in the \z{MethodTable} of the type returned from the recursion.

For both field references and method calls, we also include an implicit \z{this}
if the qualified name was in an instance context and an explicit \z{this} was
omitted.

We have several important utility methods such as \z{IsCastable},
\z{IsAssignable}, \z{IsReferenceWidening}, and so forth. These methods can be
found in \z{types/typechecker\_utils.cpp}.

\subsubsection{Constant Propagation}\label{subsubsec:const-prop}

In this phase, we rewrite the AST to contain a new type of \z{Expr}, named
\z{ConstExpr}. \z{ConstExpr} is used to mark expressions that can be resolved
at compile time. For instance, the expression `$1 + 1$' would initially be
rewritten with both of the `1's as \z{ConstExpr}s. We then have a further rule
that the addition of two \z{ConstExpr}s is a \z{ConstExpr}. In this way, we
resolve constant subtrees upwards, until we reach an unresolvable node, like a
reference to a local variable.

The code for this can be found in
\z{types/constant\_folding\_visitor.\{h,cpp\}}.  Note that this implementation
is incomplete, and only implements a subset of the behaviour specified in the
\ac{jls}. We plan to complete this implementation before starting work on code
generation, but we found that the subset implemented was sufficient to satisfy
the Assignment 4 Marmoset tests.

\subsubsection{Data-flow Analysis}\label{subsubsec:data-flow}

This phase performs another walk over the AST to check that all non-void
methods return a value, and that all code is reachable. These checks are
implemented as yet another Visitor, which can be found in
\z{types/dataflow\_visitor.\{h,cpp\}}.

The visitor recurses through all methods, maintaining a single flag named
\z{reachable}. This flag is set to \z{false} on a return statement, and an error is
emitted if a statement is visited when this flag is \z{false}. An error is also
emitted if the flag is still \z{true} after visiting the body of a method with
a non-void return type.

The visitor implements special rules for loop statements with constant
conditional expressions. Specifically, it looks for the \z{ConstExpr} nodes
emitted by the previous pass. See the \ac{jls} for these specific reachability
rules.

In addition, this phase verifies that field initializers do not reference
fields declared later in the source file.

\section{Files}

Here is a mapping from various files in the \z{types} directory to the 
phases above.

\begin{itemize}
  \item \z{types/constant\_folding\_visitor.\{h,cpp\}}: See \hyperref[subsubsec:const-prop]{Phase 4}
  \item \z{types/dataflow\_visitor.\{h,cpp\}}: See \hyperref[subsubsec:data-flow]{Phase 5}
  \item \z{types/decl\_resolver.\{h,cpp\}}: See \hyperref[subsubsec:fields]{Phase 2}
  \item \z{types/symbol\_table.\{h,cpp\}}: See \hyperref[subsubsec:type-checking]{Phase 3}
  \item \z{types/type\_info\_map.\{h,cpp\}}: See \hyperref[subsubsec:fields]{Phase 2}
  \item \z{types/typechecker.\{h,cpp\}}: See \hyperref[subsubsec:type-checking]{Phase 3}
  \item \z{types/typechecker\_errors.cpp}: See \hyperref[subsubsec:type-checking]{Phase 3}
  \item \z{types/typechecker\_utils.cpp}: See \hyperref[subsubsec:type-checking]{Phase 3}
  \item \z{types/typeset.\{h,cpp\}}: See \hyperref[subsubsec:col-types]{Phase 1}
\end{itemize}

\section{Challenges}

\subsection{Suppressing cascading errors}

When emitting errors, we put significant effort into ensuring that we would not
emit cascading errors. For instance, if we were to find a method on a type
that breaks one of the member access modifier rules, then we attempt to
suppress any errors related to its usage by blacklisting that method name. This
made our code much more complicated as we had to have reasonable behaviour for
slightly malformed programs. Specifically, passes could \emph{not} assume that
all previous passes had well-formed data. We ran into situations where the
state from previous passes was contradictory and inconsistent. These situations
were hard to debug, and led to us making extensive use of an \z{assert}-like
macro that failed much more gracefully. In the end, this helped us improve our
overall code quality.

If were to repeat this exercise, we would avoid the ad-hoc blacklisting
approach, and instead design a structured mechanism for suppressing related
errors. Indeed, we did design such a mechanism; we wanted to associate a set of
tags with each error, and deduplicate errors by tag before printing them.
However, we didn't want to retrofit this mechanism into our current compiler as
we were worried about strange interactions between this mechanism and our
current blacklisting approach.

\subsection{Specification complexity}

The \ac{jls} was an invaluable source of determining how our compiler should
behave in various interesting situations. However, mapping from a strange test
case to all relevant sections of the \ac{jls} was not trivial, to put it
mildly.

Since the \ac{jls} does not always provide algorithmic steps for determining
output, we had to reverse-engineer algorithms from prose. This often led to
incomplete algorithms when we sometimes did not consider all consequences of
the specification's lawyer-like phrasing. This was somewhat exacerbated by the
sometimes unclear mapping from Joos restrictions to \ac{jls} restrictions.

\section{Testing}

We tried to continue with our practice of writing unit tests for our various
components. In some cases, this proved effective; for instance, much of the
type checker has extensive unit tests.

However, we found that some of our tests intrinsically depend on multiple
modules. For instance, type checking variable declarations or method calls
require building a number of other data structures, \z{SymbolTable} and
\z{MethodTable} respectively. These structures themselves depend on other data
structures, like \z{TypeSet} and \z{TypeInfoMap}.

We considered importing a mocking framework, such as \z{gmock}. However, we
were not convinced that this would be a good use of our time. We would need to
write a large number of mocking helpers to make the setup portion of our tests
short.

We compromised, and wrote integration tests that covered each special case in
the type checker and other associated classes. Similar to the Marmoset tests,
the integration tests would automatically include the standard library and the
test source code.

We also included all the Marmoset tests in our repository, and ran them to
ensure we didn't introduce any regressions. One downside of these regression
tests is that they are an order of magnitude slower than unit tests. This is
explained by the fact that they need to lex, parse, weed, and type check the
standard library for every single test. We plan to investigate caching the
compilation of the standard library to decrease this high cost. Even so, the
speed of our compiler allows us to run over 1300 tests in less than 1 minute.

\end{document}

